Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-12-05 thesis/thesis.tex
thesis: Extend State of the Art chapter

\usepackage{apacite}
\usepackage{caption}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{picture}
\usepackage{subcaption}
\usepackage[nottoc]{tocbibind}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/hr_diagram_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/ngc_6494/hr_diagram_ngc_6494.png}
    \end{subfigure}
  \end{subfigure}
  \caption{Examples of typical profiles in H-R diagrams for members of open clusters. On the left Melotte 22, on the right NGC 6494.}
  \label{fig:examples_of_hr_diagrams}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/parallax_zoom_melotte_22.png}
      \caption{Melotte 22 Parallax}
      \label{fig:melotte_22_parallax_zoom}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/pm_melotte_22.png}
      \caption{Melotte 22 Proper motion}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 Parallax histogram, clearly differentiated around the value $\approx 7.3mas$ and the corresponding diagram from the photometric magnitudes.}
  \label{fig:melotte_22_pm_parallax}
\end{figure}

In many other cases things are not so simple, and the characterization process can turn very difficult and tedious using tools such as the available ones at
the Virtual Observatory (VO), and frequently it requires parametrizations based on prior knowledge about the studied region. The aim of this work is to test
an unsupervised and non-parametrized model based on Machine Learning (ML) tools.

@@

% TODO: Hablar de la misión Gaia
% TODO: Hablar de KMedias y DEC
% TODO: Mencionar trabajos relacionados

An initial approach to find OC is the search of overdensities along the galactic disk. In general, this is the initial point,
but even it looks simple, it presents a fundamental problem. The near field around the OC is fulled by two differenciated
kind of stars populations: the ones which belong to the OC (from a tens or hundreds to a few thousand) and a background composed
by thousands or millions of stars which don't belong to the OC. Find out which stars belong to the first group is the problem
treated in this work. This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, etc.).

Sometimes, this problem is easy to solve by looking into astrometric parameters, such as the analysis of proper motion of those
stars belonging to the overdensity field.

@@

Figure \ref{fig:pm_melotte_22} shows the proper motion distribution in right ascension and declination.
It is easy to see that a subgroup can be located at [20, -45].
Figure \ref{fig:pm_vec_melotte_22} shows the parallax density and confirms an overdensity at $\approx 7.5 mas$ corresponding to Mellote 22 OC.

However, as shown in Figures \ref{fig:pm_ngc_6494} and \ref{fig:parallax_ngc_6494}, in general it is not as easy and
becomes necessary to consider other parameters such as distances, or even metallicity and age (derived from isochrone curves)
and requires photometric data from stars in the studied field
\footnote{All figures have been sampled in order to reduce the document size, but are still a faithful representation of the original data}.

@@

At this point, it looks reasonable that the study of each individual cluster requires its own parametrization and technique
to identify it.

With Gaia DR2, which contains great quality astrometric and photometric data for a huge number of stars,
a new opportunity to simplify and optimize this processes is presented.
In this context new approaches are being developed to improve current detection methods and automate the OC characterization.
This work tries to take the best from \emph{Clusterix 2.0. A Virtual Observatory tool to estimate cluster membership probability}
\cite{balaguer2020clusterix} and \emph{Hunting for open clusters in Gaia DR2: 582 new open clusters
in the Galactic disc} \cite{castro2020hunting}, and merge them into a single \emph{machine learning}
tool capable of finding open clusters efficiently without requiring powerful machines to perform the computations.

The first one is a non-parametrized method. It determines empirically the frequency functions associated to astrometrical variables
without making any assumption about their profiles. However, it heavily does depend on the initial field selection and other
parameters related to frequency functions. \emph{Clusterix 2.0.} is a web app tool inside the Spanish Virtual Observatory
(VO) framework and works together with other VO tool such as:
\emph{TOPCAT} \cite{taylor2005topcat},
\emph{Aladin} \cite{bonnarel2000aladin},
\emph{VOSA} \cite{bayo2008vosa}, among other.

The second tool is based on Big Data techniques. Its purpose is to analyze the whole Gaia DR2 dataset looking for OCs by using
machine learning algorithms divided in two stages. In the first stage, the algorithm looks for posible OC candidates by searching
overdensities in the galactic disc.
The second stage studies candidates found in the first one by generating images with H-R diagrams and passing them
as sources to an Artificial Neural Network (ANN) which identifies isochrone patterns in photometric data (color and magnitude).
Before this approach was put into practice there were 1200 OCs known. Now, this number has been increased up to more than 2000 OCs
available at the Vizier catalogue \cite{ochsenbein2000vizier}.

The method developed in this project takes as base the approach made by \citeauthor{castro2020hunting} but rapidly diverges.
In first place, performance has been taken into account since the beginning, trying to get a reliable method without compromising the
needs of supercomputers.
To achieve this, instead of blindly analyzing the whole galactic disk a fields selection is made focusing in those regions with known OCs.
Then, these fields are analyzed with an homogeneous methodology based on machine learning algorithms using as sources astrometric and
photometric variables. The goal is the characterization of each individual star in the cluster, discerning wether the star belongs the OC
or not.

Thus, the main goal of this project is not the discovery of new open clusters or even candidates, but the development of a model that,
applied to already known open clusters, allows us the characterization of those OCs on an unsupervised way, with a reasonable computing
power and taking into account the same selection of astrometric and photometric variables for every OC.

% TODO: Concluir con una última sección de resumen de conclusiones, resumiendo las principales averiguaciones del estudio del dominio
% y cómo van a afectar al desarrollo específico del trabajo.

>>>>

\usepackage{apacite}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{picture}
\usepackage{subcaption}
\usepackage[nottoc]{tocbibind}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/hr_diagram_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/ngc_2682/hr_diagram_ngc_2682.png}
    \end{subfigure}
  \end{subfigure}
  \caption{Examples of typical profiles in H-R diagrams for members of open clusters. On the left Melotte 22, on the right NGC 2682.}
  \label{fig:examples_of_hr_diagrams}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/parallax_zoom_melotte_22.png}
      \caption{Melotte 22 Parallax}
      \label{fig:melotte_22_parallax_zoom}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/hr_diagram_melotte_22.png}
      \caption{Melotte 22 H-R diagram}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 Parallax histogram, clearly differentiated around the value $\approx 7.3mas$ and the corresponding diagram from the photometric magnitudes.}
  \label{fig:melotte_22_pm_parallax}
\end{figure}

In many other cases things are not so simple, and the characterization process can turn very difficult and tedious using tools such as the available ones at
the Virtual Observatory (VO), and frequently it requires parametrization based on prior knowledge about the studied region. The aim of this work is to test
an unsupervised and non-parametrized model based on Machine Learning (ML) tools.

@@

% TODO: Hablar de KMedias y DEC

An initial approach for finding OCs is the search of overdensities along the galactic disk. In general, this is the starting point but,
although it seems simple, it presents a fundamental problem already discussed. The near field around the OC is filled with two types of
distinct stars populations: those who belong to the OC (tens or hundreds to a few thousand) and a background made up of thousands or
millions of stars that are not. Finding out which stars belong to the first group is the problem treated in this work.
This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, etc.).

Sometimes, the problem is easy to solve, as we have seen, by studying astrometric parameters and looking for overdensities in the proper
motion configuration space as well as in the parallax space.

@@

Figure \ref{fig:pm_melotte_22} shows the proper motion distribution in right ascension and declination.
It is easy to see that a subgroup can be located at [20, -45].
Figure \ref{fig:pm_vec_melotte_22} shows the parallax distribution and confirms an overdensity at
$\approx 7.3 mas$ corresponding to Mellote 22 OC (See also Figure \ref{fig:melotte_22_parallax_zoom}).

However, as shown in Figures \ref{fig:pm_ngc_6494} and \ref{fig:parallax_ngc_6494}, in general it is not as easy and
becomes necessary to consider other parameters such as distances, or even metallicity and age (derived from isochrone curves)
and requires photometric data for the stars within the studied field.

@@

At this point, it seems reasonable that the study of each individual group requires its own parametrization and technique
to identify it. With Gaia DR2, which contains great high-quality astrometric and photometric data for a huge number of stars,
a new opportunity presents itself to simplify and optimize these processes. In this context, new approaches are being developed
to improve current detection methods and automate OC characterization.

\section{Gaia Mission}

In this work we make use of Gaia DR2 since, although Release 3 (Gaia DR3) is imminent, it has not been available in time for us.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission (located at L2, 1.5 million kilometers from Earth) and
operational since 2014. The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.
80\% of Gaia DR2 sources are weaker than $G_{mag} \approx 18$. For magnitudes $G \approx 21$ nominal uncertainties reach
$2 mas$ for parallaxes and $5 mas \cdot years^{-1}$ in proper motions. But for closer sources, (at the bright end, $G < 14$),
the precision is of the order of $0.02 mas$ and $0.05 mas \cdot year^{-1}$ respectively \cite{cantat2018gaia}. In general,
open clusters, which is the main subject of this work, are at optimal data thresholds.

In this context, and taking advantage of the opportunity provided by Gaia DR2, new approaches are being developed to improve current
detection methods and automating OC characterization as much as possible. This is the case of tools such as Clusterix 2.0
\cite{balaguer2020clusterix}, TOPCAT \cite{taylor2005topcat}, and VOSA \cite{bayo2008vosa}, (all of them developed by the Virtual Observatory).
These tools work together and offer good results, although it is not possible to use them in an unattended process and free of certain
different parametrization in each case.

\section{Current Methods}

TOPCAT alone does not offer an independent discrimination mechanisms to researchers. They should try an initial selection, staring from the
map of proper motions, by delimiting the area where the cluster presumably appears. When this is not possible, it is necessary to have a
certain knowledge about the cluster profile and to parametrize some approximated values. In both cases, the definition of the selected
set must then be optimized step by step until reaching a result that can be considered reasonable.

Another tool is Clusterix 2.0, which is an interactive web-based tool \cite{balaguer2020clusterix}. It takes the proper motion diagram
without assuming any prior profiles assumptions for the stars candidates membership and determines empirically the frequency functions.

Clusterix uses normal Gaussian kernel functions, defined as:

\begin{equation}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ -\frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{i} \right)^{2}}{ h^{2}} \right]}
\end{equation}

$(a, b)$ are referred to the proper motion configuration space, while a point located in the center of the cell $(i, j)$ provides the maximum
contribution for calculating the local density. In this context, the number $h$ is called the \emph{smoothing} parameter and it is measured in
the same units as proper motion.

With this same purpose of applying non-parametric methods, various clustering algorithms have been applied in the past, UPMASK \cite{krone2014upmask}
was developed to only use photometry and positions, and then it was adapted to the Gaia data \cite{cantat2018gaia} based only on proper motion and
parallax. A similar approach has been carried out using DBSCAN and machine learning algorithms to discover new clusters.

In this sense, Clusterix also claims to be a non-parametric method, but it critically depends on the initial selection for the field sizes to be analyzed.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/clusterix/selection.pdf}
  \caption{Typical Clusterix 2.0 control panel.}
  \label{fig:clusterix_control_panel}
\end{figure}

Clusterix relies on several assumptions, among other:

\begin{displayquote}
The non-field population does not occupy the entire workspace, but is spatially concentrated, which makes it possible to distinguish two regions
in the workspace: the only field region (label 'f'), dominated by stars field, and the cluster + field region (label 'c+f'), which includes field
stars and not field stars. \cite{balaguer2020clusterix}
\end{displayquote}

This fact implies defining three areas or regions with different radius. The first 'c+f' corresponds to the one in which the cluster members are
presumed to be contained together with other field stars that are not part of the cluster. The second, broader, assumes that it only contains
stars in an extended visual field without components of the cluster, and a third intermediate region, which is out of analysis (void area), which
would correspond to a possible transition zone between the other two.

The correct choice of these radii, even having a previous estimation for the 'c+f' region, significantly affects the execution of the algorithm and,
in general, requires a generously wide field 'f'. There is no general rule that defines relative proportions of these areas.

Finally, when an acceptable result is obtained, what Clusterix provides is a probability field associated with the set of stars under study.
The dataset, with an extra column for the probability value of each object to belong to the open cluster, can be exported to TOPCAT
for further optimization and final processing. Figure \ref{fig:clusterix_probability} shows an example of the probability distribution in the proper
motion space for NGC 2682.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/clusterix/statistics.pdf}
  \caption{Probability distribution in the proper motion space provided by Clusterix for NGC 2682.}
  \label{fig:clusterix_probability}
\end{figure}

Finally, it is necessary to refer to the work of \cite{castro2020hunting}, which presents a method based on machine learning techniques to make
a systematic search for overdensities in the astrometric space of the galactic disk and a subsequent identification of OC using photometric
information, also from Gaia DR2.

The method includes two phases: the first one uses an unsupervised clustering algorithm, DBSCAN, to search for overdensities
$(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$, ans subsequently applies a deep learning ANN trained with magnitude diagrams to identify isochrone
patterns within the detected overdensities and thus proceed to confirm them as OC.

It should be noted that for the execution of this method, MareNostrum 42 (Bacelona Supercomputing Center) was used. So the neural network could handle
the image recognition process with isochrone patterns and not applying theoretical models derived from values such as metallicity, masses, among other.

The result of this work is the recognition of 582 new open clusters distributed along the galactic disk for a galactic declination $b < 20^{\circ}$,
which has meant an increase of 45\% over previously ones.

Our aim in this work, linked to computational limitations compared to MareNostrum, is not to do a blind search for clusters, but to obtain a method,
in the machine learning environment, that allows the characterization of open clusters through unsupervised and non-parametrized procedures in a
single step. This method should not require further refinements with other tools.

The result of Clusterix + TOPCAT, applied to the same clusters, will be used only for comparing and validating our results.
